Safe Memory Regions for Big Data Processing

### 2016-10-19 paper/type-inference.tex
constraint syntax

\paragraph{Syntax of Constraints.}
The constraints are expressed using
a set $\regionConstants$ of region constants,
a set $\regionVars$ of region variables,
and
a set $\predVars$ of predicate variables.
% 
% ($\pi$ and $\rho$) and predicate variables ($\varphi$).
% $(\regionConstants, \regionVars, \predVars, \regionDeltaMap, \predDeltaMap, \constraintSet)$,

@@

% \emph{Region identifiers} are of two kinds: \emph{region constants} and \emph{region variables}.
Recall that a \emph{region variable} is introduced to represent an unknown actual region parameter
of a method invocation or object allocation, which will be bound to a region constant by the end of the
type inference.
A region constant may be 
(a) a \emph{formal region parameter} of a class or method, or
(b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct.

Every predicate variable $\varphi$ denotes an unknown \emph{region-constraint},
over a set of fixed formal region parameters.
%% where the set of region-constraints $\phi$ is defined by:
%% \begin{smathpar}
%% \begin{array}{lcl}
%% \phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \phi \conj \phi \\
%% \end{array}
%% \end{smathpar}
%% We will use $\rho_1 = \rho_2$ as shorthand for $\rho_1 \outlives \rho_2 \conj \rho_2 \outlives \rho_1$.
%% We will also use a set $S$ of primitive constraints of the form $\rho_1 \outlives \rho_2$
%% as shorthand for $\conj_{\phi \in S} \phi$.
%% 
We will use the term \emph{validity constraint} to denote an entailment constraint
of the form $\isvalid{\varphi}{\rho_1 \outlives \rho_2}$.

@@

The constraints generated are of the following kinds:
\begin{itemize}

\item Well-formedness constraints of form $\rho \in \rhoenv$,
restricting the domain of unification for a region variable ($\rho$)
to a constant set $\rhoenv = \{ \pi_1, \cdots, \pi_n \}$ of regions in scope,

\item Well-formedness constraints of form $\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
variable ($\varphi$) to the set of all possible constraint formulas over a fixed set of 
regions ($\rhoenv = \{ \pi_1, \cdots, \pi_n \}$) in scope, and

\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{\varphi_{\C{B}}}$
  (e.g., to indicate that the region-constraint of a derived class \C{C} is stricter than
  the region-constraint of its base class \C{B}).

\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{{F(\varphi_{\C{T}})}}$
  (e.g., to indicate the region-constraint of a class \C{C} must imply the
  region-constraint of each of its fields).

\item Validity constraints of form $\isvalid{\varphi_i \conj \phictxt} {\phicstr}$
where $\varphi_i$ is a predicate variable (representing the precondition of a
method to be determined), $\phicstr$ is a region constraint that is \emph{required}
to hold at a particular program point (within the method), and $\phictxt$ is
a region constraint that is \emph{known} to hold at that program point.

%
%% Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
%% of predicate variables and pending substitutions. While $\phictxt$
%% captures relationships that are \emph{known} to hold between concrete
%% region identifiers (i.e., $\rgn$'s) when the constraint was generated,
%% $\phicstr$ captures relationships that are \emph{required} to hold
%% among region varibles (i.e., $\rho$'s), or relationships between
%% region variables and identifiers.
%

\item Validity constraints of the form $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$
generated by an invocation of a method with precondition $\varphi_j$ (where $\phictxt$ and
$\varphi_i$ are as above).

\end{itemize}

@@

Fig.~\ref{fig:constraint-gen} illustrates this for selected language constructs.
These rules use the same context as the corresponding typing judgment in
Fig.~\ref{fig:fb-staticsem}, except that this is generalized to permit the
use of region variables and predicate variables.
Symbols $\A$ and $\env$ retain their meaning, modulo this extension.
The algorithm proceeds top-down, analyzes an expression $e$ in a context
$\stdcontext$, and returns a type $\tau$ and a set of constraints $C$
(expressed in the rules as $\exprok {\stdcontext} {e} {\tau} {C}$ ),
indicating that the expression will have a type $\tau$ provided the constraints $C$ hold.

@@

\TODO{Change the static semantics to be consistent with this, and
make this a comment there.}
One notable aspect here is the following. Consider a method with predicate
variable $\varphi_m$ in a class with predicate variable $\varphi_c$. 
We use $\varphi_m$ as the constraint formula in the initial context used for
the method (instead of $\varphi_m \conj \varphi_c$ as in the static semantics).
It is possible to show that there is no loss of generality in doing so.
Specifically, it can be shown for any well-typed program, replacing the constraint
$\varphi_m$ of every method by the constraint $\varphi_m \conj \varphi_c$,
where $\varphi_c$ is the constraint of the class containing the method,
produces a well-typed program.

We use the above simplification to avoid generating validity constraints of the form
$\isvalid{\varphi_m \conj \varphi_c \conj \phi_{cx}}{\phi_{cs}}$ (i.e., constraints
with more than one predicate variable on the left hand side, which introduce complications).
%  ensures that the generated constraints produces a restricted form of constraints that is easier to solve.

>>>>

\paragraph{Syntax of Constraints.}
(See~Fig.~\ref{fig:constraint-syntax}.)
The constraints are expressed using a set $\regionConstants$ of region constants,
a set $\regionVars$ of region variables, and a set $\predVars$ of predicate variables.

@@

Recall that a \emph{region variable} is introduced to represent an unknown actual region parameter
of a method invocation or object allocation, which will be bound to a region constant by the end of the
type inference.
A region constant may be 
(a) a \emph{formal region parameter} of a class or method, or
(b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct.

A predicate variable $\varphi$ is introduced to represent, \eg, the unknown precondition of a method.
The constraint-solver will end up binding it to a \emph{region-constraint} $\phi$
over a set of fixed formal region parameters.

%% where the set of region-constraints $\phi$ is defined by:
%% \begin{smathpar}
%% \begin{array}{lcl}
%% \phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \phi \conj \phi \\
%% \end{array}
%% \end{smathpar}
%% We will use $\rho_1 = \rho_2$ as shorthand for $\rho_1 \outlives \rho_2 \conj \rho_2 \outlives \rho_1$.
%% We will also use a set $S$ of primitive constraints of the form $\rho_1 \outlives \rho_2$
%% as shorthand for $\conj_{\phi \in S} \phi$.
%% 
% We will use the term \emph{validity constraint} to denote an entailment constraint
% of the form $\isvalid{\varphi}{\rho_1 \outlives \rho_2}$.

@@

The constraints generated are primarily of the form $\isvalid{\varphi_i \conj \phictxt} {\phicstr}$
or $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$.  Here, $\varphi_i$ is a predicate variable
(representing the precondition of a method to be determined), $\phicstr$ is a region-constraint
that is \emph{required} to hold at a particular program point (within the method), and $\phictxt$ is
a region-constraint that is \emph{known} to hold at that program point.
Validity constraints of the form $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$
are generated by an invocation of a method with precondition $\varphi_j$.

We use \emph{well-formedness constraints} of the form $\rho \in \rhoenv$
to restrict the domain of unification for a region variable ($\rho$)
to a constant set $\rhoenv = \{ \pi_1, \cdots, \pi_n \}$ of regions in scope,
and  well-formedness constraints of form $\tywf{\rhoenv}{\varphi}$ to
restrict the domain of a predicate variable ($\varphi$) to the set of all possible
constraint formulas over a fixed set of  regions ($\rhoenv = \{ \pi_1, \cdots, \pi_n \}$) in scope.

%The constraints generated are of the following kinds:
%\begin{itemize}
%
%\item Well-formedness constraints of form $\rho \in \rhoenv$,
%restricting the domain of unification for a region variable ($\rho$)
%to a constant set $\rhoenv = \{ \pi_1, \cdots, \pi_n \}$ of regions in scope,
%
%\item Well-formedness constraints of form $\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
%variable ($\varphi$) to the set of all possible constraint formulas over a fixed set of 
%regions ($\rhoenv = \{ \pi_1, \cdots, \pi_n \}$) in scope, and
%
%\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{\varphi_{\C{B}}}$
%  (e.g., to indicate that the region-constraint of a derived class \C{C} is stricter than
%  the region-constraint of its base class \C{B}).
%
%\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{{F(\varphi_{\C{T}})}}$
%  (e.g., to indicate the region-constraint of a class \C{C} must imply the
%  region-constraint of each of its fields).
%
%\item Validity constraints of form $\isvalid{\varphi_i \conj \phictxt} {\phicstr}$
%where $\varphi_i$ is a predicate variable (representing the precondition of a
%method to be determined), $\phicstr$ is a region constraint that is \emph{required}
%to hold at a particular program point (within the method), and $\phictxt$ is
%a region constraint that is \emph{known} to hold at that program point.
%
%\item Validity constraints of the form $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$
%generated by an invocation of a method with precondition $\varphi_j$ (where $\phictxt$ and
%$\varphi_i$ are as above).
%
%\end{itemize}

@@

Fig.~\ref{fig:constraint-gen-0} illustrates this for selected language constructs.
These rules use the same context as the corresponding typing judgment in
Fig.~\ref{fig:fb-staticsem}, except that this is generalized to permit the
use of region variables and predicate variables.
Symbols $\A$ and $\env$ retain their meaning, modulo this extension.
The algorithm proceeds top-down, analyzes an expression $e$ in a context
$\stdcontext$, and returns a type $\tau$ and a set of constraints $C$
(expressed in the rules as $\exprok {\stdcontext} {e} {\tau} {C}$ ),
indicating that the expression will have a type $\tau$ provided the constraints $C$ hold.

@@

%\TODO{Change the static semantics to be consistent with this, and
%make this a comment there.}
%One notable aspect here is the following. Consider a method with predicate
%variable $\varphi_m$ in a class with predicate variable $\varphi_c$. 
%We use $\varphi_m$ as the constraint formula in the initial context used for
%the method (instead of $\varphi_m \conj \varphi_c$ as in the static semantics).
%It is possible to show that there is no loss of generality in doing so.
%Specifically, it can be shown for any well-typed program, replacing the constraint
%$\varphi_m$ of every method by the constraint $\varphi_m \conj \varphi_c$,
%where $\varphi_c$ is the constraint of the class containing the method,
%produces a well-typed program.
%
%We use the above simplification to avoid generating validity constraints of the form
%$\isvalid{\varphi_m \conj \varphi_c \conj \phi_{cx}}{\phi_{cs}}$ (i.e., constraints
%with more than one predicate variable on the left hand side, which introduce complications).
%%  ensures that the generated constraints produces a restricted form of constraints that is easier to solve.
